Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-28 Dail_Mad_ext.tex
likelihood

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$).
%%%% I agree with your comment, but I think we need to distinguish
%%%% between things like actual and apparent survival probs.
%\hl{[Feels a little repetitious with the material
%before the equation, although I understand you're describing S and G
%there and omega and gamma here.  In general, these sections
%describing the DM models are clear and well written, but we could
%possibly spend less time on them b/c it's already published.
%Possibly it's better this way, though.]}
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_{it} \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is the sum %\hl{[I think!]}
of the birth rate and the immigration rate,
i.e. the recruitment rate. In cases where no movement occurs, the DM model can directly estimate
demographich parameters, and this mechanistic description of
population change is much more appealing than classical
population growth models such as the logistic model. However, movement
complicates the interpretation of the the parameters, reducing the
connection to specific population processes. In such cases, the model
of dynamics shown in Eq~\ref{eq:Nt} can be replaced with standard
population models as described in Sect~\ref{sec:ext}. Irrespective of
the model for population dynamics, it is worthwhile to note that the
discrete distributions used in the DM model avoid the issues
associated with the Gaussian distribution used to model process
variation in standard state-space models.

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
Surprsingly detection probability, $p$, can be estimated without
secondary sampling occasions. However, this is only possible because
if the parameteric assumptions of the population dynamics are met
exactly. In practice, it is preferable to obtain direct information
about $p$ using secondary sampling occassions or some other auxiliary
data, as is demonstrated in section

@@

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation.

@@

Here is some BUGS code: \hl{[This is code for a model we haven't
presented yet.  Could we move these sections after model extensions?]}
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    mu[i,t-1] = N[i,t-1]*exp(r)
    N[i,t] ~ dpois(mu[i,t-1])
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

@@

>>>>

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_{it} \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is the sum %\hl{[I think!]}
of the birth rate and the immigration rate,
i.e. the recruitment rate. In the former case where no movement
occurs, the DM model can directly estimate
demographich parameters, and this mechanistic description of
population change is much more appealing than classical
population growth models such as the logistic model. However, movement
complicates the interpretation of the the parameters, reducing the
connection to specific population processes. In such cases, the model
of dynamics shown in Eq.~\ref{eq:Nt} can be replaced with standard
population models as described in Sect~\ref{sec:ext}. Irrespective of
the model for population dynamics, it is worthwhile to note that the
discrete distributions used in the DM model avoid the issues
associated with the Gaussian distribution used to model process
variation in standard state-space models.

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection, and the one
considered in the original formulation, is
\begin{equation}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
Surprsingly detection probability, $p$, can be estimated without
secondary sampling occasions. However, this is only possible
if the parameteric assumptions of the population dynamics are met
exactly. In practice, it is preferable to obtain direct information
about $p$ using secondary sampling occassions or some other auxiliary
data, \hl{as is demonstrated in section....}

@@

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation. For Markovian models, this can be done recursively, making
the likelihood tractible. A general expression for the likelihood of
the DM model is:
\begin{equation}
  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{ \prod_{t=T}^2 \left\{ \sum_{N_{it}=0}^{\infty}
  [n_{it}|N_{it}][N_{it}|N_{it-1}] \right\} \right\} \right\}
  \label{eq:like}
\end{equation}

@@

Here is some BUGS code:

\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma)
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

@@

\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    mu[i,t-1] = N[i,t-1]*exp(r)
    N[i,t] ~ dpois(mu[i,t-1])
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}
